\scshape Evolutionarily stable strategy analysis and its links to demography and genetics through invasion fitness

### 2023-01-30 rev1_cover_letter.tex
just need final edits

I would like to thank the Editor and Reviewers for their very detailed in insightful comments. I have copied each Reviewer's comments below followed by a detailed response with changes I've made in response to the comments.

@@

I've had added that the fitness $w(x,y)$ is a linear function of its arguments, which are the phenotypes of the two interacting individuals. Linearity here is reasonable given the phenotype is a probability of selecting among two pure strategies.

@@

I have reworked those sentences and have added a list of the assumptions after Result 1 with some discussion of how they might be relaxed.

@@

Yes, I have largely left out reviewing adaptive dynamics, which is not directly intentional per se but more because I wanted to build around the ESS condition in the original 1973 paper as the Phil Trans issue commemorates 50 years since that paper. However, adaptive dynamics is of course extremely important to evolutionary game theory and to the concept of invasion fitness. I've added XXX...
%%% Add something about convergence stability to discussion.

As to whether invasion fitness in adaptive dynamics is more general than what I discuss here, I would guess the reviewer is thinking of cases where short-term evolution reaches a non-equilibrium attractor like a cycle or chaotic attractor and invasion fitness must then really be measured in the sense of a Lyapunov exponent. I agree those cases are important but I do not mention them so as to keep the scope of the review from growing too much. It should be noted too that the assumptions and tools employed in typical deterministic and stochastic adaptive dynamics models ignore the non-equilibrium cases too as they are difficult to handle in general. One of those assumptions is often, as I discuss in reference to long-term evolution, infrequent mutation, which when combined with finite population size, leads to at most two genotypes existing at a time (see the Champagnat 2006 ref) and one of them either going extinct or fixing before the arrival of the next mutation.

Finally, I totally agree that the second ESS condition from the 1973 paper is really the convergence stability condition. My review wasn't structured to synthesize the original ESS conditions with the stability and convergence conditions from adaptive dynamics but this would be the basis from a great review or synthesis article. %% Add something here maybe if I write convergence stability paragraph?

@@

I think the reviewer may be referring to Maynard Smith (1978, Ann Rev Ecol Syst), Grafen (1979, Anim Behav), and Hines and Maynard Smith (1979, JTB).

@@

I agree that I need to be clearer about the conditions necessary the results in sections 3 and 4 to hold. I've now added a bulleted list at the end of section 4 that lists all of the conditions and also discuss how restrictive they may or may not be.

@@

Yes, this limitation is important. I now specify it clearly as the first assumption in the bulleted list at the end of section 4.

@@

%% mention this more clearly at the end of section 4 and/or in the discussion

@@

%% ugh. ok take care of this

@@

I agree that section 6 is more technical. I have added XXX..
%% add here

I have tried to keep Section 7 to a minimum of technical depth. I do not agree that its understandable to only technically sophisticated readers. However, it is relevant to add some of the assumptions for the modifier models discussed and I now do that in XXXX
%% add here

@@

I mention the reduction principle since it uses external stability to show the general evolutionary principle that selection in a constant generates purifying selection for the highest fitness genotype and against genetic variation created by other genotypes. Varying environments don't exhibit the reduction principle since genetic or phenotypic variation can be adaptive in these cases. I don't see how discussing technical criticisms of the reduction principle advances the purpose of the section, which is to show how the ESS concept as it relates to external stability can study the evolution of genetic and demographic parameters in the case of varying environments.

>>>>

I would like to thank the Editor and Reviewers for their very detailed in insightful comments. I have copied each Reviewer's comments below (in blue) followed by a detailed response with changes I've made in response to the comments.

@@

I've had added that the fitness $w(x,y)$ is a linear function of its arguments, which are the phenotypes of the two interacting individuals. Linearity here is reasonable given the phenotype is a probability of selecting among two pure strategies. I provide more discussion of this assumption in the new section 8.

@@

I have reworked those sentences. I have also added a whole new section 8 that details all of the assumptions for Result 1 and discusses them in turn. See some of my additional responses below to Reviewer 2 for more detail.

@@

Yes, I had largely left out reviewing adaptive dynamics, which is not directly intentional per se but more because I wanted to build around the ESS condition in the original 1973 paper as the Phil Trans issue commemorates 50 years since that paper. However, adaptive dynamics is of course extremely important to evolutionary game theory and to the concept of invasion fitness. I've have added a new section 9 with a short discussion of adaptive dynamics and evolutionary branching.

As to whether invasion fitness in adaptive dynamics is more general than what I discuss here, I would guess the reviewer is thinking of cases where short-term evolution reaches a non-equilibrium attractor like a cycle or chaotic attractor and invasion fitness must then really be measured in the sense of a Lyapunov exponent. I agree those cases are important but I do not mention them to limit the length of the review. It should be noted too that the assumptions and tools employed in typical deterministic and stochastic adaptive dynamics models ignore the non-equilibrium cases too as they are difficult to handle in general. One of those assumptions is often, as I discuss in reference to long-term evolution, infrequent mutation, which when combined with finite population size, leads to at most two genotypes existing at a time (see the Champagnat 2006 ref) and one of them either going extinct or fixing before the arrival of the next mutation.

Finally, I totally agree that the second ESS condition from the 1973 paper is really the convergence stability condition. My review wasn't structured to synthesize the original ESS conditions with the stability and convergence conditions from adaptive dynamics but this would be the basis from a great review or synthesis article. However, I now briefly mention in section 9 how the second condition connects to convergence stability.

@@

I think the reviewer may be referring to Maynard Smith (1978, Ann Rev Ecol Syst), Grafen (1979, Anim Behav), and Hines and Maynard Smith (1979, JTB). While this history is interesting, I haven't added a discussion of those papers since there isn't room in this paper for a proper analysis of the relevant issues.

@@

I agree that I need to be clearer about the conditions necessary the results in sections 3 and 4 to hold. I've now added a new section 8 with a bulleted list that lists all of the conditions. The new section then discusses these assumptions in some detail.

@@

Yes, this limitation is important. I now specify it clearly as the first assumption in the bulleted list at in the new section 8. I also mention the rock paper scissors example.

@@

I now cite Friedman in addition to the rock paper scissors model as examples where ESS may not imply stability.

@@

This is very interesting and I would like to learn more but I could not locate where in the Karlin and Lessard book they clearly described these simulations and the reviewer does not note whether they have published their own simulations. Based on what I think is a general lack of published clear simulation tests of long-term evolution, I think that what the reviewer is describing is a fruitful area for future research that isn't mature enough for me to cover here.

@@

I agree that section 6 is more technical. Nevertheless, I hope that it still gives the reader a sense for how invasion fitness can be used in these situations. On the other hand, I have tried to keep Section 7 to a minimum of technical depth. I do not agree that its understandable to only technically sophisticated readers. With respect to the assumptions that of the modifier models, they share a number in common with the broader long-term evolution approach and those assumptions are now discussed in detail in section 8.
%% add here

@@

I mention the reduction principle since it uses external stability to show the general evolutionary principle that selection in a constant environment generates purifying selection for the highest fitness genotype and against genetic variation created by other genotypes. Varying environments don't exhibit the reduction principle since genetic or phenotypic variation can be adaptive in these cases. I don't see how discussing technical criticisms of the reduction principle advances the purpose of the section, which is to show how the ESS concept as it relates to external stability can study the evolution of genetic and demographic parameters in the case of varying environments.
